---
title: "DE violin plots"
author: "Madeline Iseminger"
date: "06/04/2021"
output: github_document
---
# Set up libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# load packages
library(Seurat)
library(patchwork)
library(ggplot2)
library(cowplot)
library(dplyr)
```
Run using Ubuntu 20.04, R 4.0.5, RStudio Version 1.4.1103. 

# Load data, name clusters by cell type, and create subsets

Load the binary datafile of filtered and integrated data.
```{r}
dt <- readRDS("~/Desktop/AllSamplesMergedFiltered/AllSamplesMergedFiltered")
```

Prepare cluster ID to cell type conversion:
```{r}
# get conversion for cluster names and cell types (this is from Trevor's code)
cluster_ID_list <- list(Neutrophils = c("0", "1", "4", "5", "9", "10", "11"),
                        Monocytes = c("2", "8"), 
                        Macrophages = c("3", "7"), 
                        T_cells_NK_cells = "6", 
                        Epithelial_cells = "12", 
                        B_cells = "13")

# to call cluster labels, use code like: cluster_ID_list["Neutrophils"] or cluster_ID_list$Neutrophils

# merge clusters with the same cell type by renaming clusters
new.cluster.ids <- c("Neutrophils", "Neutrophils", "Monocytes", "Macrophages", "Neutrophils", "Neutrophils", "T_cells_NK_cells", "Macrophages", "Monocytes", "Neutrophils", "Neutrophils", "Neutrophils", "Epithelial_cells", "B_cells")
names(new.cluster.ids) <- levels(dt)
nameddt <- RenameIdents(dt, new.cluster.ids) #this is the dt but with names clustered
```

Function to subset the Seurat object (this chunk from Trevor's code):

```{r}
# create function to subset the Seurat object 
subsetdt <- function(COVID_severity, tissue_type, cell_type){
  
  if (is.na(COVID_severity) == FALSE) {
  # subset by clinic status
  dt.sub <- subset(dt, clinic_status == COVID_severity)
  }
  
  if (is.na(tissue_type) == FALSE) {
  # subset by tissue type
  dt.sub <- subset(dt, tissue == tissue_type)
  }
  # subset by cell type
  if (cell_type == "Neutrophils") {
    cell_ID <- 1
  }
  if (cell_type == "Monocytes") {
    cell_ID <- 2
  }
  if (cell_type == "Macrophages") {
    cell_ID <- 3
  }
  if (cell_type == "T_cells_NK_cells") {
    cell_ID <- 4
  }
  if (cell_type == "Epithelial_cells") {
    cell_ID <- 5
  }
  if (cell_type == "B_cells") {
    cell_ID <- 6
  }
  dt.sub <- subset(dt.sub, seurat_clusters == cluster_ID_list[[cell_ID]])
  # create df
  Idents(dt.sub) <- c("male", "female")
  return(dt.sub)
}
```

  
Subset the Seurat object by disease severity and sample type for plotting later:
```{r eval=FALSE}
#mild disease, B cells
mildBcell <- subsetdt(COVID_severity = "Mild COVID", tissue_type = NA, cell_type = "B_cells")

#mild disease, epithelial cells
mildepicell <- subsetdt(COVID_severity = "Mild COVID", tissue_type = NA, cell_type = "Epithelial_cells")

#mild disease, monocytes
mildMono <- subsetdt(COVID_severity = "Mild COVID", tissue_type = NA, cell_type = "Monocytes")

#mild severity, clusters 0-13 to work with Trevor's subset() function
mild <- subset(dt, clinic_status == "Mild COVID")

#mild severity, clusters grouped by name
namedmild <- subset(nameddt, clinic_status == "Mild COVID")
```

# Find genes to test

Find names of genes of interest present in the data.  Genes that would be interesting to search for include:

- ACE2: present are "BACE2" "ACE2" 
- IFIT2
- IFIT3
- IL18: present are "IL18R1"  "IL18RAP" "IL18BP"  "IL18"
- TLR3 and TLR7: endosomal RNA receptors that recognize SARS-CoV-2 RNA and begin the immune response cascade
- IFN genes: type I interferons with antiviral effects; from https://pubmed.ncbi.nlm.nih.gov/33071044/, type I IFN (IFN-I) should be more highly expressed in females. Present genes containing "IFN" are "IFNGR1" "IFNAR2" "IFNAR1" "IFNGR2" "IFNLR1" "IFNG" "IFNG-AS1" "IFNL1"

```{r}
# get list of gene names over all samples and cell types
gene_names <- rownames(dt@assays$RNA@counts)

# check if a gene is present in the data
gene = "IFN" # string to search for in array of gene names
gene_names[grep("CD200R", gene_names)] # prints a list of gene names containing that string
```

# Make plots 
Make violin plots for individual genes:
```{r}
# ACE2 in B cells, mild disease
VlnPlot(mildBcell, features = c("ACE2"))
VlnPlot(mildBcell, features = c("IFIT2"))

# ACE2 with cell type names
VlnPlot(nameddt, features = c("ACE2"), split.by = "sex")

# ACE2 with cell type names, mild disease only
VlnPlot(namedmild, features = c("ACE2"), split.by = "sex")
VlnPlot(namedmild, features = c("IFIT2"), split.by = "sex")
VlnPlot(namedmild, features = c("CD200R1"), split.by = "sex")
VlnPlot(namedmild, features = c("TLR3", "TLR7"), split.by = "sex", combine = FALSE)
VlnPlot(namedmild, features = c("IL18"), split.by = "sex")

# mild disease, monocytes
VlnPlot(mildMono, features = c("CD200R1"))
VlnPlot(mildMono, features = c("TLR3", "TLR7"), split.by = "sex", combine = FALSE)
```
# Sample composition by cell type
Plot proportion of cell type in different sample types:
```{r}
# get proportion of cells per cluster in object
overall <- prop.table(table(Idents(nameddt)))

# plot bar graph of proportion of total cells from each cluster by sex (BAL only)
barplot(overall, main="Total Cells", ylab="Proportion of cells", xlab="Cell Type", las=2, main = "All Cells")

# get proportion of cells per cluster in object
mildtable <- prop.table(table(Idents(namedmild)))

# plot bar graph of proportion of total cells from each cluster by sex (BAL only)
barplot(mildtable, main="Total Cells", ylab="Proportion of cells", xlab="Cell Type", las=2, main = "Mild COVID")
# get average expression for each gene per cluster
cluster.averages <- AverageExpression(dt)
```
