---
title: "DE violin plots"
author: "Madeline Iseminger"
date: "06/04/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# clear global workspace
rm(list=ls())

# load packages
library(Seurat)
library(patchwork)
library(ggplot2)
library(cowplot)
library(dplyr)
library(enrichR) # for GO analysis

# SET UP MULTI-CORES
#no_cores <- availableCores() - 1
#future::plan(multisession, workers = no_cores)
```

## R Markdown
Goal: Differential Expression between female/male patients: control vs mild vs severe

Run using Ubuntu 20.04, R 4.0.5, RStudio Version 1.4.1103.  To install Seurat:

```{r eval=FALSE}
install.packages("Seurat")
```

Load the binary datafile of filtered and integrated data.
```{r}
myobj <- readRDS("~/Desktop/AllSamplesMergedFiltered/AllSamplesMergedFiltered")
```

Prepare cluster ID to cell type conversion, search for genes of interest that are available in this dataset:
```{r}
# get conversion for cluster names and cell types (this is from Trevor's code)
cluster_ID_list <- list(Neutrophils = c("0", "1", "4", "5", "9", "10", "11"),
                        Monocytes = c("2", "8"), 
                        Macrophages = c("3", "7"), 
                        T_cells_NK_cells = "6", 
                        Epithelial_cells = "12", 
                        B_cells = "13")

# to call cluster labels, use code like: cluster_ID_list["Neutrophils"] or cluster_ID_list$Neutrophils

# merge clusters with the same cell type by renaming clusters
new.cluster.ids <- c("Neutrophils", "Neutrophils", "Monocytes", "Macrophages", "Neutrophils", "Neutrophils", "T_cells_NK_cells", "Macrophages", "Monocytes", "Neutrophils", "Neutrophils", "Neutrophils", "Epithelial_cells", "B_cells")
names(new.cluster.ids) <- levels(myobj)
myobj <- RenameIdents(myobj, new.cluster.ids)
```

Find names of genes of interest present in the data.  Genes that would be interesting to search for include:

- ACE2: present are "BACE2" "ACE2" 
- IFIT2
- IFIT3
- IL18: present are "IL18R1"  "IL18RAP" "IL18BP"  "IL18"
- TLR3 and TLR7: endosomal RNA receptors that recognize SARS-CoV-2 RNA and begin the immune response cascade
- IFN genes: type I interferons with antiviral effects; from https://pubmed.ncbi.nlm.nih.gov/33071044/, type I IFN (IFN-I) should be more highly expressed in females. Present genes containing "IFN" are "IFNGR1" "IFNAR2" "IFNAR1" "IFNGR2" "IFNLR1" "IFNG" "IFNG-AS1" "IFNL1"
- viral genes if present in the data
- cytokines lumped together

```{r}
# get list of gene names over all samples and cell types
gene_names <- rownames(myobj@assays$RNA@counts)

# check if a gene is present in the data
gene = "IFN" # string to search for in array of gene names
gene_names[grep(gene, gene_names)] # prints a list of gene names containing that string
```

Subset the Seurat object by disease severity and sample type for plotting later:
```{r eval=FALSE}
# myobj$clinic_status == "Severe COVID" # this makes a mask of True/False

# severe is commented out because my computer breaks on it
#myobj.severe <- subset(myobj, clinic_status == "Severe COVID")
myobj.mild <- subset(myobj, clinic_status == "Mild COVID")
myobj.healthy <- subset(myobj, clinic_status == "Healthy control")

# myobj.BAL <- subset(myobj, tissue == "BAL")
# myobj.blood <- subset(myobj, tissue == "Blood")
```

Make violin plots for individual genes:
```{r}
# all individuals: genes split by sex, all cell types
VlnPlot(myobj, features = c("IFIT2", "IFIT3"), split.by = "sex", combine = FALSE)

# mild disease: genes split by sex, all cell types
plots <- VlnPlot(myobj.mild, features = c("IFIT2", "IFIT3"), split.by = "sex", 
    pt.size = 0.5, combine = FALSE)

# show plots in .md file
plots
```

Plot proportion of cell type in different sample types:
```{r}
# get proportion of cells per cluster in object
overall <- prop.table(table(Idents(myobj)))

# plot bar graph of proportion of total cells from each cluster by sex (BAL only)
barplot(overall, main="Total Cells", ylab="Proportion of cells", xlab="Cell Type", las=2)

# get average expression for each gene per cluster
cluster.averages <- AverageExpression(myobj)
```

## GO enrichment analysis from a list of genes

Fetch GO terms for selected genes and plot them with their functions.

```{r}
# select database
setEnrichrSite("Enrichr") # human genes
websiteLive <- TRUE
# list available databases
dbs <- listEnrichrDbs()

# list of genes to query
geneList <- c("IFIT2", "IFIT3")

# query database
dbs <- c("GO_Molecular_Function_2015", "GO_Cellular_Component_2015", "GO_Biological_Process_2015")
if (websiteLive) {
    enriched <- enrichr(geneList, dbs)
}

#view results table
if (websiteLive) enriched[["GO_Biological_Process_2015"]]

if (websiteLive) plotEnrich(enriched[[3]], showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value")
```
